\documentclass{beamer}
\usepackage{tikz}
\usepackage{lstautogobble, listings, xcolor} 

\definecolor{bluekeywords}{rgb}{0.13, 0.19, 0.7}
\definecolor{goldenkeywords}{rgb}{0.67, 0.58, 0.13}
\definecolor{greencomments}{rgb}{0.1, 0.5, 0.2}
\definecolor{redstrings}{rgb}{0.8, 0.15, 0.1}
\definecolor{graynumbers}{rgb}{0.5, 0.5, 0.5}
\definecolor{subtlegray}{rgb}{0.98, 0.98, 0.98}

\lstset{
    autogobble,    
    columns=fullflexible,
    showspaces=false,
    showtabs=false,
    breaklines=true,
    showstringspaces=false,
    numbers=left,
    breakatwhitespace=true,
    escapeinside={(*@}{@*)},
    rulecolor=\color{lightgray},
    backgroundcolor=\color{subtlegray},
    commentstyle=\color{greencomments},
    keywordstyle=\color{bluekeywords},
    ndkeywordstyle=\color{goldenkeywords},
    stringstyle=\color{redstrings},
    numberstyle=\color{graynumbers},
    basicstyle=\ttfamily\linespread{1.15}\footnotesize,
    frame=tb,
    framesep=12pt,
    framexleftmargin=12pt,
    tabsize=4,
    captionpos=b
}

\lstdefinelanguage{TML}{ 
    keywords={changeto, move, goto, if, switch, while, module, accept, reject, alphabet},
    ndkeywords={left, right, tapehead, blank},
    sensitive=true,
    comment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]',
    morestring=[b]"
}

\usetikzlibrary{automata, positioning, arrows, shapes}

\title{Turing Machine Language}
\author{Pete Gautam}
\institute{University of Glasgow}

\begin{document}

    \frame{\titlepage}

    \begin{frame}{Introduction}
        \begin{itemize}
            \item Turing Machines (TM) are a model of computation.
            \item They are typically taught theoretically, i.e. pen-and-paper.
            \item Students find it hard to learn the content.
            \item Perhaps it is easier to learn as a programming language (PL)?
        \end{itemize}
    \end{frame}

    \begin{frame}{Turing Machines}
        \begin{itemize}
            \item Can be thought of as boolean functions on string
            \item Has states and transitions
            \item Can be represented as a directed graph (called FSM)
        \end{itemize}

        \begin{figure}[htb]
            \centering
            \begin{tikzpicture}
                \node[state, accepting] (q0) at (0, 0) {$q_0$};
                \node[state] (q1) at (2.5, 0) {$q_1$};
                \node[state, fill=green, opacity=0.6] (A) at (5, -1) {$A$};
                \node[state, fill=red, opacity=0.6] (R) at (5, 1) {$R$};
        
                \draw[->] (q0) edge[loop above] node {$0|1, R$} (q0);
                \draw[->] (q0) -- node[above] {$\#, L$} (q1);
                \draw[->] (q1) -- node[below, rotate=-20] {$0 \to \#, L$} (A);
                \draw[->] (q1) -- node[above, rotate=20] {$1|\#, L$} (R);
            \end{tikzpicture}
            \caption{A FSM representation of TMs.}
        \end{figure}
    \end{frame}
    
    \begin{frame}{Project Overview}
        \begin{enumerate}
            \item Creating the Turing Machine Language (TML).
            \item Constructing a parser for TML.
            \item Constructing the product (website) to showcase the parser.
        \end{enumerate}
    \end{frame}

    \begin{frame}{Language}
        \lstinputlisting[language=TML]{code/isDiv2.txt}
    \end{frame}

    \begin{frame}{Parser}
        \begin{figure}[htb]
            \centering
            \begin{tikzpicture}
                \node[draw] (CW) at (0, 0) {Lexical Analysis};
                \node[draw] (CP) at (0, -1.5) {Syntactic Analysis};
                \node[draw] (CV) at (0, -3) {Semantic Analysis};
                \node[draw] (CC) at (-2, -4.5) {Code Generation};
                \node[draw] (CE) at (2, -4.5) {Code Execution};
        
                \draw[->] (0, 1) -- node[left] {Source Program} (CW);
                \draw[->] (CW) -- node[left] {Program Stream} (CP);
                \draw[->] (CP) -- node[left] {Program Context} (CV);
                \draw[->] (CV) -- node[left] {Program Context} (CC);
                \draw[->] (CV) -- node[right] {Program Context} (CE);
                \draw[->] (CC) -- node[left] {Turing Machine} (-2, -5.5);
            \end{tikzpicture}
            \caption{The parsing process.}
        \end{figure}
    \end{frame}

    \begin{frame}{Product}
        \begin{enumerate}
            \item Homepage
            \begin{enumerate}
                \item Editor
                \item Convert to FSM and definition
                \item Execute on tape
            \end{enumerate}
            \item Documentation Pages
            \item Error Pages
        \end{enumerate}
        % TODO: DEMO SHOULD BE AROUND THIS TIME
    \end{frame}

    \begin{frame}{Evaluation}
        \begin{itemize}
            \item Unit testing throughout
            \begin{itemize}
                \item Parser
                \item Website- less successful due to mocking
            \end{itemize}
            \item User evaluation
            \begin{itemize}
                \item Worksheet on TM and TML.
                \item With 18 second year CS students.
                \item It is easier to grasp TML than TM, but the results might not apply in general.
            \end{itemize}
        \end{itemize}
    \end{frame}

    \begin{frame}{Future Work}
        \begin{itemize}
            \item Improve the language with more features:
            \begin{itemize}
                \item \texttt{move end};
                \item parametrisation.
            \end{itemize}
            \item Make the website more accessible:
            \begin{itemize}
                \item Add a play button;
                \item Allow the FSM to be zoomed in.
            \end{itemize}
        \end{itemize}
    \end{frame}
    
\end{document}
